---
title: "COGS212 CourseProject - Exploratory Data Analysis (EDA)"
output: html_document
date: "2023-10-24"
---

<!-- DH general comments

- Mostly good use of literate code style, following Peng & Matsui: introducing your project, walking us through your data, identifying notable features

- Missing info on data provenance

- Lots of issues with immutability violations, long lines

- Project would benefit from organization: storing the documents you've already produced, along with the report and narrative

-->


```{r setup, include=FALSE}
## DH: I think this is only necessary if you're knitting to slides
knitr::opts_chunk$set(echo = TRUE)
```

This project explores interrelated research questions: 1) how has internal migration within the United States changed over time?  2) when people do migrate internally, where do they migrate from and 3) where do they migrate to? 

Load libraries: 
```{r}
library(here)
library(skimr)
library(tidyr)
library(stringr)
library(dplyr)
library(visdat)
library(ggplot2)
```

<!-- DH: What is this dataset?  Where did it come from?  -->
Read in data: 
```{r}
df_raw <- read.csv(here("county_migration_data.csv")) 
```

Check the packaging:
```{r}
skim(df_raw)
```

Each column is a year of data. Therefore, I need to turn the data frame from a
wide format to a long format.

First, I need to rename the columns: 
```{r}

colnames(df_raw)
## DH: Violates immutability; append this to read.csv() with a pipe
colnames(df_raw) <- c("origin", "destination", as.character(1990:2010))
colnames(df_raw)

```

I can now transform the dataset from a wide format to a long format: 
```{r}
## DH: long line; spacing around =
df_long <- gather(df_raw, 
                  year, 
                  migrants, 
                  "1990":"2010", 
                  factor_key = TRUE)
```

As a quick check: 
```{r}
21 * 164403
```
Looks like the transformation worked! 
(Note: 21 is the number of years and 164403 is the number of rows in the original dataset.) <!-- DH: Express the comparison directly: `identical(nrow(df_long), 21L * nrow(df_raw))`.  And you can make it an assertion by putting it in stopifnot(). -->

Look at the top of the dataset: 
```{r}

head(df_long)

```

Look at the bottom of the dataset: 
```{r}

tail(df_long)

```

<!-- DH: Is fips code 99999 meaningful?  Or something like a NA? -->

This dataset only provides the FIPS codes for the counties. In this project, I 
am going to focus on between state-level rather than county-level migration. The 
first two numbers indicate the state; therefore, I can use these codes to 
determine the origin state and destination state. More information on the FIPS
codes can be found here: https://transition.fcc.gov/oet/info/maps/census/fips/fips.txt. 
To do this, I need to 1) add a zero at the beginning of the FIPS code for codes 
with only four numbers, 2) separate the first two numbers from the remainder of
the FIPS code, 3) create a key that matches FIPS code to state name, and 4) merge
this key with my data frame. 

Step 1: Add zeros
```{r}
## DH: long lines; violate immutability
## DH: DRY: create a helper function and apply it to both `origin` and `destination` at the same time
df_long$origin_new <- ifelse(nchar(df_long$origin) < 5, paste("0", df_long$origin, sep = ""), df_long$origin)

df_long$destination_new <- ifelse(nchar(df_long$destination) < 5, paste("0", df_long$destination, sep = ""), df_long$destination)
```


Step 2: Separate first two numbers
```{r}
## DH: violate immutability
df_long$origin_state_num <- str_extract(df_long$origin_new, "^\\d{2}")
df_long$destination_state_num <- str_extract(df_long$destination_new, "^\\d{2}")

```

Check that these columns look correct - values should be between 01 and 56
```{r}

unique(df_long$origin_state_num)
unique(df_long$destination_state_num)

```

The 99 values are missing values (it is unknown where people came from or went)! 
I'm going to do a check check of how many observations are missing (as a percentage
of the data frame): 
```{r}

(sum(df_long$origin_state_num == "99")/nrow(df_long)) * 100
(sum(df_long$destination_state_num == "99")/nrow(df_long)) * 100

```
It is about 2% of each variable. <!-- DH: Are these values exactly the same? --> I am going to remove those now, as to ensure 
this does not create a problem when I try merging my key onto this data set:
```{r}

df_long_sub <- subset(df_long, origin_state_num != "99")
sum(df_long_sub$origin_state_num == "99")

df_long_sub <- subset(df_long_sub, destination_state_num != "99")
sum(df_long_sub$destination_state_num == "99")

```

Step 3: Create key
(Note: I am creating this key based on the documentation linked above.)
```{r}
## DH: long lines
FIPS <- c("01", "02", "04", "05", "06", "08", "09", as.character(10:13), 
          as.character(15:42), as.character(44:51), as.character(53:56))
Statename <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", 
               "Connecticut", "Delaware", "District of Columbia", "Florida", 
               "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", 
               "Kansas", "Kentucky", "Louisiana", "Maine", "Maryland", 
               "Massachusetts", "Michigan", "Minnesota", "Mississippi", "Missouri", 
               "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", 
               "New Mexico", "New York", "North Carolina", "North Dakota", "Ohio", 
               "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "South Carolina",
               "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", 
               "Washington", "West Virginia", "Wisconsin", "Wyoming")

key <- data.frame(FIPS, Statename)

key
```


Step 4: Merge key with data
First merge onto origin:
```{r}
## DH: violates immutability
colnames(key) <- c("origin_state_num", "origin_state_name")
## DH: don't introduce new dependencies in the middle of the script; note that you can do the same thing with dplyr::left_join()
df_state <- plyr::join(df_long_sub, key, by = c("origin_state_num"))

```

Then merge onto destination: 
```{r}
## DH: violates immutability
colnames(key) <- c("destination_state_num", "destination_state_name")
df_state <- plyr::join(df_state, key, by = c("destination_state_num"))

```

I am now going to subset to only the columns that I need for my analysis: 
```{r}
## DH: long line; violates immutability
df_state <- subset(df_state, select = c(origin_state_num, origin_state_name, destination_state_num, destination_state_name, year, migrants))

```

Check my "n's"
Year:
```{r}

df_state |> 
    count(year)

```

<!-- DH: All correct? -->

Origin state: 
```{r}

df_state |> 
    count(origin_state_name)

```


Destination state: 
```{r}

df_state |> 
    count(destination_state_name)

```

We can see that not all states have the same number of observations. This is
because the data still reflects the county-level data, NOT state-level. I need
to summarize the data by year and state-dyad: 
```{r}
## DH: long line; are you intentionally leaving df_state_agg grouped? 
df_state_agg <- df_state |> 
  group_by(origin_state_name, origin_state_num, destination_state_name, destination_state_num, year) |> 
  summarize(migrants = sum(migrants, na.rm = TRUE))
  
```

Let's look at the n's now: 
```{r}

df_state_agg |> 
    count(origin_state_name)

df_state_agg |> 
    count(destination_state_name)

```

Looks good now! <!-- DH: Is the length what we'd expect? -->


I am going to double check that I have no missingness in my data set: 
```{r}

vis_miss(df_state_agg)

```

I am going to create the following variables that will be useful to answering my research questions: 1) total migrants that year, 2) total migrants from a given 
origin state, and 3) total migrants to a given destination state.
```{r}
## DH: violates immutability; this could just be a single pipe
df_state_agg <- df_state_agg |>
  group_by(year) |>
  mutate(migrants_year = sum(migrants, na.rm = TRUE))

df_state_agg <- df_state_agg |>
  group_by(origin_state_name) |>
  mutate(migrants_origin = sum(migrants, na.rm = TRUE))

df_state_agg <- df_state_agg |>
  group_by(destination_state_name) |>
  mutate(migrants_destination = sum(migrants, na.rm = TRUE))

```

Before I start plotting, I need to remove sceientific notation: 
```{r}
options(scipen=999)
```

I am now going to look at the general distribution of the migrant variable: 
```{r}

plot(table(df_state_agg$migrants))

```
As expected, we see that there are no migrants in most cases. <!-- DH: Why is this expected? -->

This plot looks at how the number of migrants has changed over time: 
```{r}

ggplot(df_state_agg, aes(x = year, y = migrants_year, group = 1)) +
  geom_line()+
  geom_point()

```

Creating tables of migrants by destination  and origin state: 
```{r}
## DH: long line
destination <- subset(df_state_agg, select = c(destination_state_name, migrants_destination))
## DH: violates immutability; could've just started the pipe at the previous line
destination <- destination %>% 
  group_by(destination_state_name) %>%
  summarize(migrants_destination = sum(migrants_destination, na.rm = TRUE))

origin <- subset(df_state_agg, select = c(origin_state_name, migrants_origin))
origin <- origin %>% 
  group_by(origin_state_name) %>%
  summarize(migrants_origin = sum(migrants_origin, na.rm = TRUE))
  
```


This is a bar plot of all destination states...: 
```{r}
## DH: long line; spacing around =; flip the axes for readability
ggplot(df_state_agg, aes(x=reorder(destination_state_name, migrants_destination), y=migrants_destination)) + 
    geom_bar(stat = "identity") + 
    coord_flip() +
    labs(x = "State Name (Destination)", y = "Number of Migrants")

```

...and of top 10 destination states: 
```{r}
## DH: long lines, spacing
destination_10 <- subset(df_state_agg, destination_state_name %in% c("California", "Texas", "New York", "Florida", "Illinois", "Pennsylvania", "Ohio", "Michigan", "New Jersey", "Georgia"))

ggplot(destination_10, aes(x=reorder(destination_state_name, -migrants_destination), y=migrants_destination)) + 
  geom_bar(stat = "identity") + 
  labs(x = "State Name (Destination)", y = "Number of Migrants")

```

This is a bar plot of all origin states...:
```{r}
## DH: long lines; flip axes; spacing around =
ggplot(df_state_agg, aes(x=reorder(origin_state_name, -migrants_origin), y=migrants_origin)) + 
  geom_bar(stat = "identity") + 
  labs(x = "State Name (Origin)", y = "Number of Migrants")

```

...and of top 10 origin states: 
```{r}
## DH: same issues
origin_10 <- subset(df_state_agg, origin_state_name %in% c("California", "Texas", "New York", "Florida", "Illinois", "Pennsylvania", "Ohio", "Michigan", "New Jersey", "Georgia"))

ggplot(destination_10, aes(x=reorder(destination_state_name, -migrants_destination), y=migrants_destination)) + 
  geom_bar(stat = "identity") + 
  labs(x = "State Name (Destination)", y = "Number of Migrants")

```

The top 10 states that people migrate from are the same states that people migrate to. A follow-up question is whether this could change if I looked at migration per capita instead. Unfortunately, this data is unable to answer this question. We also see that migration has increased between 1990 and 2010. This data is sufficient to answer my three original research questions. <!-- DH: What are the answers? -->
